### 2014-01-19 ace-pldi14/rebuttal.txt

We thank the reviewers for their detailed and insightful comments!

Due to the unusual number of reviews, we hope the committee does not mind that we exceeded the 500 word soft limit proportionally. To save time, the items most relevant to each reviewer are summarized below.

R1 - 1, 5, 6
R2 - 1, 5, 6, 9
R3 - 1, 2, 5, 6, 7
R4 - 2, 5, 6
R5 - 2, 3, 4, 5, 6, 8

Corrections:
1. R1 describes Ace as providing "syntax-directed" mechanisms, which would indeed be "routine". This is incorrect -- Ace uses a _type-directed_ mechanism to determine the semantics of an expression. That is, there is _not_ a direct mapping from syntax to semantics; the mapping depends on logic associated with the type of a designated subexpression. This represents an inversion of control relative to syntax-directed mechanisms. As R2 and R3 suggest by their characterizations of it, this mechanism is the main novel technical innovation Ace introduces.

2. R5 states that Ace's types are not "types in a conventional sense at all". This is incorrect -- Ace's types are constructs that statically classify expressions and are erased at run-time. There is no "dynamic typechecking" of Ace function bodies. Our theoretical work [1] connects our mechanism to type-level computation, further emphasizing that active types are clearly static types (this paper also contains a much simpler example, which R4 may be interested in). We will take care to emphasize this in the final version. Note to R3: the word "form" should be used for the syntactic form (e.g. Attribute); type is reserved for the static type.

3. R5 states that "Lisp-style macros are compositional" in the sense that we use the word. Figure 1 did in fact state this (though there are subtleties, see 5 below). Note  that macros do not permit non-local type system extensions.

4. R5 is incorrect that "all of the related work covered in Figure 1 is compositional". Systems like Qi/Shen, Typed Racket and Xroma do not prevent the introduction of conflicting rules that could cause problems by giving a type to constructs that another extension wished to classify as ill-typed or by violating previously maintained invariants (e.g. by violating the representational consistency check that Ace uses, see Sec. 3.1.3).

Clarifications:
5. We do agree with the reviewers that our explanation of extension composition in Ace could have been clearer, particularly because safe extensibility is a primary goal of Ace. There are two facets that we handle: "non-interference" -- each extension can claim exclusive control over a set of expressions, and "compositional reasoning" -- each extension can assume that expressions of another type are representationally consistent (have a consistent type in the target language, as determined by the trans_type method, cf. Sec. Sec. 3.1.3), making it possible to reason modularly about well-typedness of the translation. Both hinge on our novel type-directed dispatch protocol and our use of active types. We will take care to clarify this distinction. A fully-rigorous treatment of this topic requires a type theoretic formulation beyond the scope of this paper. Such a treatment can be found in [1].

6. Several reviewers (R1, R3, R5) had questions about binding and scope. We concur  that this could be discussed in more detail. As stated in Sec. 3.2, the active base controls local variable binding and sets up the typing context. The context can be structured as the base sees fit, but it is typically a dictionary mapping variable names to (active) types. Active types do not access it directly; recursing on a subexpression that contains variables will eventually delegate to the active base to extract type information.

6.1. Active types can add new contexts other than typing contexts to the "context" object by adding a new attribute to it upon first use of a construct that requires it. There is no guarantee that other types cannot manipulate this context, but if a globally unique attribute name is used, this will not occur accidentally (see 7 below) -- the new context will be statically threaded through the program unchanged until the next construct that intentionally accesses it is encountered. We hope to pursue a cleaner foundation for custom contexts in future work.

6.2. References to variables that are not local, like the names of other functions in scope, can also be handled by the active base by reference to the Python environment that the function is defined within. Ace captures this when a generic function is defined (we inadvertently did not state this). The special Python attribute ace_type designates an Ace type for an object in the Python environment. That type must determine how to lift this value from compile-time to run-time given the active target. This is how the OpenCL bindings in 2.6 work -- the ace_type attribute of OpenCL buffer objects (e.g. d_input) is an instance of clx.TypedBuffer parameterized by run-time type information available from the numpy object that was used to create the buffer (e.g. the numpy dtype attribute). R3 is correct that this interactive mode of use implies an "intermingling" of Python's dynamic semantics and Ace's static semantics -- this can be seen as a form of staged compilation (and we will describe it more clearly as such in Sec. 2.6).

6.3. Handling of nested function definitions (and lambdas) is also delegated entirely to the active base. For example, an active base supporting a C target could support a nested OpenCL-based function by using the OpenCL host API internally to compile it.

7. R3 asks for a description of a "well-behaved extension". These are extensions that do not use Python's metaprogramming facilities to modify other active types/bases or core Ace data structures directly, and that do not attempt to modify the base's typing context or contexts defined by other types. We will discuss this alongside our discussion of composability in the final version.

8. R5 states that LMS is the "system most similar" to Ace. Ace permits type system extensions while LMS uses Scala's monolithic (albeit expressive) type system directly. It is thus only comparable to the translation functionality Ace provides. We cited Delite but not LMS under "cross-compilation". We will cite both in the final version and discuss this distinction further in the text.

9. We are sympathetic to worries that the examples in Sec. 4 are not fully detailed. This is due to the 10-page limit. We will decrease the number of non-code examples to increase the level of detail for the ones with code in the final version. R2 suggests that we first develop more experience with Ace. We emphasize that we have developed a large number of extensions, including a full implementation of an industrial language -- OpenCL. Not discussed in this paper is a large distributed scientific simulation framework running on clusters of GPUs that has been used to conduct neurobiological circuit simulations written atop Ace (see Sec. XX of [2]). Several course and undergraduate research projects have now used Ace as well (e.g. [3]).

[1] TODO: ESOP draft
[2] TODO: SC draft
[3] TODO: Nathan's thesis

### 2014-01-21 ace-pldi14/rebuttal.txt

We thank the reviewers for their detailed and insightful comments! We would like to correct some errors that appear to have been made in the reviews and provide requested clarifications. Due to the unusual number of reviews, we hope the committee does not mind that we have exceeded the 500 word soft limit proportionally. The items most relevant to each reviewer are summarized below.

@@

R1 - 1, 5, 6
R2 - 1, 5, 6, 9
R3 - 1, 2, 5, 6, 7
R4 - 2, 5, 6
R5 - 2, 3, 4, 5, 6, 8

Corrections:
1. R1 describes Ace as providing "syntax-directed" mechanisms, which would indeed be "routine". This is incorrect -- Ace uses a _type-directed_ mechanism to determine the semantics of an expression. That is, there is _not_ a direct mapping from syntax to semantics; the mapping depends on logic associated with the type of a designated subexpression. This represents an inversion of control relative to conventional syntax-directed mechanisms. As R2 and R3 suggest by their characterizations of it, this is the main novel technical innovation Ace introduces.

2. R5 states that Ace's types are not "types in a conventional sense at all". This is incorrect -- Ace's types are constructs that statically classify expressions and are erased at run-time. There is no "dynamic typechecking" of Ace function bodies. Our theoretical work [1] connects our mechanism to type-level computation, further emphasizing that active types are clearly static types. This paper also contains a much simpler example, as requested by R4. We will take care to emphasize this in the final version.

Note to R3: we use the word "form" for the syntactic form (e.g. Attribute, BinOp); "type" is reserved for static types.

@@

3. R5 states that "Lisp-style macros are compositional" in the sense that we use the word. Figure 1 did in fact state this (though there are subtleties, see 5 below). Note that macros do not permit non-local type system extensions.

4. R5 is incorrect that "all of the related work covered in Figure 1 is compositional". Systems like Qi/Shen, Typed Racket and Xroma allow extension writers to add to a "bag of rules" and so do not prevent the introduction of conflicting rules that could cause problems by, e.g., giving a type to constructs that another extension wished to classify as ill-typed or by violating previously maintained invariants (e.g. by violating the representational consistency check that Ace uses, see Sec. 3.1.3).

Clarifications:
5. We do agree with the reviewers that our explanation of extension composition in Ace could have been clearer and made in one place in the paper. There are two facets that we handle: 
  a) "non-interference" -- each extension has _exclusive_ control over a set of expressions, per the dispatch protocol
  b) "compositional reasoning" -- each extension can assume that expressions of another type are representationally consistent (have a consistent type in the target language, as determined by the trans_type method, cf. Sec. Sec. 3.1.3), making it possible to reason modularly about well-typedness of the translation.

Both hinge on our novel type-directed dispatch protocol and our use of active types. We will take care to clarify this issue. A fully-rigorous treatment of this topic requires a type theoretic formulation beyond the scope of this paper. Such a treatment can be found in [1].

@@

6. Several reviewers (R1, R3, R5) had questions about binding and scope. We concur that this could be discussed in more detail. As stated in Sec. 3.2, the active base controls local variable binding and sets up the typing context. The context can be structured as the base sees fit, but it is typically a dictionary mapping variable names to (active) types. Active types do not access it directly; recursing on a subexpression that contains variables will eventually delegate to the active base to extract type information.

6.1. Active types can add new contexts other than typing contexts to the "context" object by adding a new attribute to it upon first use of a construct that requires it. There is no guarantee that other types cannot manipulate this context, but if a globally unique attribute name is used, this will not occur accidentally (see 7 below); the new context will be statically threaded through the program unchanged until the next construct that intentionally accesses it is encountered. We hope to pursue a cleaner foundation for custom contexts in future work.

6.2. References to variables that are not local, like the names of other functions in scope, can also be handled by the active base by reference to the Python environment that the function is defined within. Ace captures this when a generic function is defined (we inadvertently did not state this).

The special Python attribute ace_type can be used to designate an Ace type for an object in the Python environment. That type must determine how to lift this value from compile-time to run-time given the active target. This is how the OpenCL bindings in 2.6 work -- the ace_type attribute of OpenCL buffer objects (e.g. d_input) is an instance of clx.TypedBuffer parameterized by run-time type information available from the numpy object that was used to create the buffer (e.g. the numpy dtype attribute). R3 is correct that this interactive mode of use implies an "intermingling" of Python's dynamic semantics and Ace's static semantics -- this can be seen as a form of staged compilation (and we will describe it more clearly as such in Sec. 2.6).

@@

6.3. Handling of nested function definitions (and lambdas) is also delegated entirely to the active base. For example, an active base supporting a C target could support a nested OpenCL-based function by using the OpenCL host API internally to compile it.

7. R3 asks for a description of a "well-behaved extension". These are extensions that do not use Python's metaprogramming facilities to modify other active types/bases or core Ace data structures directly, and that do not attempt to modify the base's typing context or contexts defined by other extensions. We will discuss this alongside our discussion of composability in the final version.

@@

8. R5 states that LMS is the "system most similar" to Ace. Ace permits type system extensions while LMS uses Scala's monolithic (albeit expressive) type system directly. It is thus only comparable to the translation functionality Ace provides. We cited Delite but not LMS under "cross-compilation". We will cite both in the final version and discuss this distinction further in the text.

9. We are sympathetic to worries that the examples in Sec. 4 are not fully detailed. This is due to the 10-page limit. We will decrease the number of non-code examples to increase the level of detail for the ones with code in the final version. R2 suggests that we first develop more experience with Ace. We emphasize that we have developed a large number of extensions, including a full implementation of an industrial language -- OpenCL -- using Ace. In addition, though not discussed in this paper, we have developed a large distributed scientific simulation framework running on clusters of GPUs that has been used to conduct neurobiological circuit simulations (see Sec. 5 of [2]). Finally, several course and undergraduate research projects have now used Ace as well (e.g. [3]).

[1] https://github.com/cyrus-/papers/blob/master/esop14/att-esop14.pdf?raw=true
[2] https://github.com/cyrus-/papers/blob/master/sc13/ace.pdf?raw=true
[3] http://src.acm.org/2013/NathanFulton.pdf

### 2014-02-06 thesis-proposal/acc.tex
More progress on proposal

In all of these systems, the code completion interface has remained primarily menu-based. When an item is selected, code is inserted immediately, without further input from the developer. These systems are also difficult to extend: a fixed strategy determines the completions that are available, so library providers cannot directly specify new domain-specific or contextually-relevant logic. 
In this paper we propose a technique called {\it active code completion} that eliminates these restrictions\footnote{Portions of this work previously appeared in a poster abstract \cite{ACC_VLHCC}.}. This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler. This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.

>>>>

In all of these systems, the code completion interface has remained primarily menu-based. When an item is selected, code is inserted immediately, without further input from the developer. These systems are also difficult to extend: a fixed strategy determines the completions that are available, so library providers cannot directly specify new domain-specific or contextually-relevant logic. 
In this paper we propose a technique called {\it active code completion} that eliminates these restrictions. This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler. This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.

### 2014-02-06 thesis-proposal/background.tex
More progress on proposal

% !TEX root = omar-thesis-proposal.tex
\subsubsection{Background}\label{alibs}
The term \emph{active libraries} was first introduced by Veldhuizen et al. \cite{activelibraries, active-libraries-thesis} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''. The authors suggested a number of reasons libraries might benefit by being able to influence the programming system at compile-time or edit-time, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler).

% This stuff isn't actually hooking into the compiler...
The first concrete realizations of active libraries, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation. A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{veldhuizen2000blitz++}. Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about (see discussion in \cite{Robison:2001:IEC:376656.376751}). %They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.

More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking and reporting\- logic and extending a language with new abstractions. These mechanisms are highly expressive because they allow users to programmatically manipulate syntax trees directly, but they suffer from problems of composability and safety. For example, compile-time macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all the code that they enclose. This can be problematic, however, as outer macros can interfere with the functionality of inner macros. Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a value in the underlying language (a problem fundamentally related to the compatibility problems described in Section \ref{motivation}). Thus, macros can be used to automate code generation, but not to truly extend the semantics of a language. It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.

Some term rewriting systems replace the block scoping of macros with pattern-based dispatch. Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting passes into the compiler from within libraries \cite{activelibraries}. Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows libraries to define custom compile-time term rewriting logic if an appropriate flag is passed in \cite{jones2001playing}. In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms are highly expressive and avoid some of the difficulties of block-scoped macros. But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently. It is also difficult to guarantee that such logic is correct and difficult to reason about code when importing a library can change the semantics of the program in a non-local manner.
% http://www.haskell.org/ghc/docs/latest/html/users_guide/rewrite-rules.html

Another example of an active library approach to extensibility is SugarJ \cite{erdweg2011sugarj} and other generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}. These languages permit libraries to extend the base syntax of the core language in a nearly-arbitrary manner, and these extensions are imported transitively throughout a program. These extensions are also not safely composable. For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms. If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same standard notations but back them with differing implementations. 
And again, it can be difficult for users to predict what type of data an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.

>>>>

% !TEX root = omar-thesis-proposal.tex
\subsubsection{Background}\label{alibs}\label{background}
The term \emph{active libraries} was first introduced by Veldhuizen et al. \cite{activelibraries, active-libraries-thesis} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''. The authors suggested a number of reasons libraries might benefit by being able to influence the programming system at compile-time or edit-time, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler).

% This stuff isn't actually hooking into the compiler...
The first concrete realizations of active libraries in statically typed settings, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation. A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{veldhuizen2000blitz++}. Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about (see discussion in \cite{Robison:2001:IEC:376656.376751}). %They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.

More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and extending a language with new abstractions. These mechanisms are highly expressive because they allow users to programmatically manipulate syntax trees directly, but they suffer from problems of composability and safety. For example, compile-time macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all of the code that they enclose. This can be problematic, however, as outer macros can interfere with the functionality of inner macros. Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a value in the underlying language (a problem fundamentally related to the problem of relying on a common intermediate language,  described in Section \ref{external-approaches}). Thus, macros can be used to automate code generation, but not to truly extend the semantics of a language. It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.

Some term rewriting systems replace the explicitly delimited scoping of macros with global pattern-based dispatch. Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting passes into the compiler from within libraries \cite{activelibraries}. Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows providers to introduce custom compile-time term rewriting logic if an appropriate flag is passed in \cite{jones2001playing}. In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms are highly expressive and avoid some of the difficulties of explicitly invoked macros. But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently. It is also difficult to guarantee that such logic is correct and difficult to reason about code when simply importing a library can change the semantics of the program in a highly non-local manner.
% http://www.haskell.org/ghc/docs/latest/html/users_guide/rewrite-rules.html

Another example of an active library approach to extensibility is SugarJ \cite{erdweg2011sugarj} and other languages generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}. These languages permit libraries to extend the base syntax of the core language in a nearly arbitrary manner, and these extensions are imported transitively throughout a program. Unfortunately, this flexibility means that  extensions are also not safely composable. For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms. If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same notations but back them with differing implementations. 
And again, it is difficult to predict what an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.
